### 2022-08-01 paper/localpsf.tex
draft before taking into account Noemi's comments
remove verb font from tables
edited abstract based on Noemi's comments

@@

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. Such operators arise as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in PDE constrained inverse problems. We compute impulse responses of the operator at a collection of scattered points, then interpolate translated versions of these impulse responses to approximate entries of the integral kernel. The interpolation uses a new approximation we call ``local mean displacement invariance,'' which improves upon classical local translation invariance.
%	Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. 
	Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many point sources as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, and these moments are computed for impulse responses at all points simultaneously via a procedure that involves applying the operator to a small number of polynomial functions.
	The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Fast H-matrix methods are used to perform further matrix computations in nearly-linear complexity, including matrix-matrix addition and multiplication, matrix inversion and factorization, and matrix-vector products.
	We use the method to approximate the Hessian in an ice sheet flow PDE constrained inverse problem. Numerical results demonstrate that our method substantially outperforms existing methods. We are able to approximate high rank operators using a small number of operator applies.
\end{abstract}

@@

\begin{table}
	{\small
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{3pt}
			\renewcommand{\arraystretch}{1.1}
			\begin{tabular}{c| c c c | c c c | c c c}
				&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
				\hline
				Iter & 
				\#CG & \#Stokes & $||\bm{g}||$ & 
				\#CG & \#Stokes & $||\bm{g}||$ & 
				\#CG & \#Stokes & $||\bm{g}||$ \\
				\verb~0~ &
				\verb~1~ & \verb~4~ & \verb~2.0e+7~ &
				\verb~1~ & \verb~4~ & \verb~2.0e+7~ &
				\verb~1~ & \verb~4~ & \verb~2.0e+7~ \\
				\verb~1~ &
				\verb~2~ & \verb~6~ & \verb~5.8e+6~ &
				\verb~2~ & \verb~6~ & \verb~7.9e+6~ &
				\verb~2~ & \verb~6~ & \verb~5.8e+6~ \\
				\verb~2~ &
				\verb~4~ & \verb~10~ & \verb~2.4e+6~ &
				\verb~5~ & \verb~12~ & \verb~3.9e+6~ &
				\verb~4~ & \verb~10~ & \verb~2.4e+6~ \\
				\verb~3~ &
				\verb~2~ & \verb~6+22~ & \verb~5.8e+5~ &
				\verb~13~ & \verb~28~ & \verb~1.6e+6~ &
				\verb~13~ & \verb~28~ & \verb~5.8e+5~ \\
				\verb~4~ &
				\verb~5~ & \verb~12~ & \verb~5.6e+4~ &
				\verb~42~ & \verb~86~ & \verb~4.8e+5~ &
				\verb~33~ & \verb~68~ & \verb~9.1e+4~ \\
				\verb~5~ &
				\verb~10~ & \verb~22~ & \verb~3.5e+3~ &
				\verb~84~ & \verb~170~ & \verb~7.7e+4~ &
				\verb~57~ & \verb~116~ & \verb~6.2e+3~ \\
				\verb~6~ &
				\verb~14~ & \verb~30~ & \verb~2.7e+1~ &
				\verb~125~ & \verb~252~ & \verb~5.0e+3~ &
				\verb~76~ & \verb~154~ & \verb~1.1e+2~ \\
				\verb~7~ &
				\verb~0~ & \verb~2~ & \verb~4.1e-2~ &
				\verb~194~ & \verb~390~ & \verb~7.9e+1~ &
				\verb~117~ & \verb~236~ & \verb~2.3e-1~ \\
				\verb~8~ & 
				\verb~---~ & \verb~---~ & \verb~---~ &
				\verb~0~ & \verb~2~ & \verb~1.6e-1~ &
				\verb~0~ & \verb~2~ & \verb~4.1e-2~ \\
				\hline
				Total & 
				\verb~38~ & \verb~114~ & \verb~---~ &
				\verb~466~ & \verb~950~ & \verb~---~ &
				\verb~303~ & \verb~624~ & \verb~---~ \\
			\end{tabular}
			\endgroup
		\end{center}
	}
	\caption{Convergence history for solving the Stokes inverse problem using preconditioned inexact Newton PCG. Preconditioners shown are our method with five impulse response batches (PSF (5)) constructed at the third iteration, regularization preconditioning (REG), and no preconditioning (NONE). Columns titled \#CG show the number of PCG iterations used to solve the Newton system for $\bm{\widehat{\basalfriction}}_k$. Columns titled $||\bm{g}||$ show the $l^2$ norm of the gradient at $\bm{\basalfriction}_k$. Columns titled \#Stokes show the total number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves for $u$ performed during the linesearch going from $\bm{\basalfriction}_{k-1}$ to $\bm{\basalfriction}_k$, plus one adjoint Stokes solve to compute the gradient at $\bm{\basalfriction}_k$, plus one incremental forward and one incremental adjoint Stokes solve per PCG iteration for solving the Newton system. In the PSF (5) portion of row 3, we write $6+22$ to indicate that 6 Stokes solves were used during the standard course of the iteration, and 22 Stokes solves were used to build the PSF (5) preconditioner. }
	\label{tab:newton_convergence_table}
\end{table}

@@

\begin{figure}
	\begin{subfigure}{0.59\textwidth}
		\begin{center}
			\begin{tikzpicture}[baseline, trim axis right]
			\begin{axis}[width=0.99\linewidth, height=0.8\linewidth,
			ymode=log,
			xmin=-50,
			xmax=1450,
			ymax=1.e4,
			xlabel={$k$, generalized eigenvalue \#},
			ylabel={$\lambda_{k}$},
			xtick={0, 300, 600, 900, 1200},
			%legend pos=north east,
			legend style={font=\small},
			ylabel near ticks, ylabel shift={-6pt},
			axis x line*=bottom,
			axis y line*=left,
			legend style={at={(0.95,1)},anchor=north east},
			title={Generalized eigenvalues, $\bm{H} \bm{u}_{k}=\lambda_{k} \bm{\preconditioner} \bm{u}_{k}$}
			]
			\addlegendentry{REG}
			\addplot[dotted, line width=0.5mm] table[x=k, y=reg] {generalizedEigenvalues.dat};
			\addlegendentry{PSF (1)}
			\addplot[blue, line width=0.5mm] table[x=k, y=psf1] {generalizedEigenvalues.dat};
			\addlegendentry{PSF (5)}
			\addplot[red, line width=0.5mm] table[x=k, y=psf5] {generalizedEigenvalues.dat};
			\addlegendentry{PSF (25)}
			\addplot[purple, line width=0.5mm] table[x=k, y=psf25] {generalizedEigenvalues.dat};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{}
		\label{fig:generalized_eigenvalues}
	\end{subfigure}
	\begin{subfigure}{0.39\textwidth}
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{4pt}
			\renewcommand{\arraystretch}{1.25}
			\begin{tabular}{c| c c}
				& $\kappa(\bm{\preconditioner}^{-1} \bm{H})$ &	$\kappa(\bm{\preconditioner}^{-1} \bm{H}_{\text{gn}})$ \\
				\hline 
				REG        & \verb~7.02e+3~ & \verb~7.02e+3~ \\
				NONE       & \verb~8.71e+2~ & \verb~9.22e+2~ \\
				PSF $(1)$  & \verb~1.75e+1~ & \verb~1.76e+1~ \\
				PSF $(5)$  & \verb~1.39e+1~ & \verb~1.39e+1~ \\
				PSF $(25)$ & \verb~1.03e+1~ & \verb~1.06e+1~
			\end{tabular}
			\endgroup
		\end{center}
		\caption{}
		\label{fig:condition_number}
	\end{subfigure}
	\caption{Left (\ref{fig:generalized_eigenvalues}): eigenvalues for generalized eigenvalue problem $\bm{H} \bm{u}_k = \lambda_k \bm{\preconditioner} \bm{u}_k$. Here $\bm{H}$ is the Hessian and $\bm{\preconditioner}$ is the PSF approximation constructed using the Gauss-Newton Hessian, $\bm{H}_\text{gn}$, for 1, 5, and 25 batches (PSF (1), PSF (5), and PSF (25), respectively), or the regularization Hessian (REG). Right (\ref{fig:condition_number}): condition number, $\kappa$, for $\bm{\preconditioner}^{-1} \bm{H}$ and $\bm{\preconditioner}^{-1} \bm{H}_\text{gn}$ for these different preconditioners, and no preconditioner (NONE). All operators are evaluated at the $\bm{\basalfriction}$ that solves the inverse problem for $\gamma=7.3 \times 10^3$, as determined by the Morozov discrepancy principle.}
\end{figure}

>>>>

% CITATIONS TO DO
% - Steffen Börm, Lars Grasedyck and Wolfgang Hackbusch:
%"Introduction to Hierarchical Matrices with Applications",
%Engineering Analysis with Boundary Elements 27(5), pp. 405--422, 2003.
%- Lars Grasedyck, Ronald Kriemann and Sabine Le Borne:
%"Parallel blackbox H-LU preconditioning for elliptic boundary value problems",
%Computing and Visualization in Sciences, Vol. 11, pp 273--291, 2007.
%- Ronald Kriemann:
%"H-LU Factorization on Many-Core Systems",
%Computing and Visualization in Science, 16, pp. 105-117, 2015.

@@

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF)-based method for approximating operators that have locally supported non-negative integral kernels. Such operators arise as Schur complements when solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in PDE-constrained inverse problems. The proposed method uses impulse responses of the underlying operator (computed at a collection of scattered points) and interpolates translated versions of these impulse responses to approximate entries of the integral kernel. The interpolation uses a ``local mean displacement invariance'' approximation, which improves upon classical local translation invariance. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many point sources as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, and these moments are computed for impulse responses at all points simultaneously via a procedure that involves applying the operator to a small number of polynomial functions. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Fast H-matrix methods are used to perform further matrix computations in nearly-linear complexity, including matrix-matrix addition and multiplication, matrix inversion and factorization, and matrix-vector products. We illustrate our approach by applying the method to approximate the Hessian in an ice sheet flow PDE constrained inverse problem. Numerical results demonstrate that the proposed method substantially outperforms existing methods. We show that high rank operators can be efficiently approximated using only a small number of operator applies.
\end{abstract}

@@

\begin{table}
	{\small
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{3pt}
			\renewcommand{\arraystretch}{1.1}
			\begin{tabular}{c| c c c | c c c | c c c}
				&  \multicolumn{3}{|c|}{PSF (5)} & \multicolumn{3}{|c|}{REG} & \multicolumn{3}{|c}{NONE} \\
				\hline
				Iter & 
				\#CG & \#Stokes & $||\bm{g}||$ & 
				\#CG & \#Stokes & $||\bm{g}||$ & 
				\#CG & \#Stokes & $||\bm{g}||$ \\
				0 &
				1 & 4 & 2.0e+7 &
				1 & 4 & 2.0e+7 &
				1 & 4 & 2.0e+7 \\
				1 &
				2 & 6 & 5.8e+6 &
				2 & 6 & 7.9e+6 &
				2 & 6 & 5.8e+6 \\
				2 &
				4 & 10 & 2.4e+6 &
				5 & 12 & 3.9e+6 &
				4 & 10 & 2.4e+6 \\
				3 &
				2 & 6+22 & 5.8e+5 &
				13 & 28 & 1.6e+6 &
				13 & 28 & 5.8e+5 \\
				4 &
				5 & 12 & 5.6e+4 &
				42 & 86 & 4.8e+5 &
				33 & 68 & 9.1e+4 \\
				5 &
				10 & 22 & 3.5e+3 &
				84 & 170 & 7.7e+4 &
				57 & 116 & 6.2e+3 \\
				6 &
				14 & 30 & 2.7e+1 &
				125 & 252 & 5.0e+3 &
				76 & 154 & 1.1e+2 \\
				7 &
				0 & 2 & 4.1e-2 &
				194 & 390 & 7.9e+1 &
				117 & 236 & 2.3e-1 \\
				8 & 
				--- & --- & --- &
				0 & 2 & 1.6e-1 &
				0 & 2 & 4.1e-2 \\
				\hline
				Total & 
				38 & 114 & --- &
				466 & 950 & --- &
				303 & 624 & --- \\
			\end{tabular}
			\endgroup
		\end{center}
	}
	\caption{Convergence history for solving the Stokes inverse problem using preconditioned inexact Newton PCG. Preconditioners shown are our method with five impulse response batches (PSF (5)) constructed at the third iteration, regularization preconditioning (REG), and no preconditioning (NONE). Columns titled \#CG show the number of PCG iterations used to solve the Newton system for $\bm{\widehat{\basalfriction}}_k$. Columns titled $||\bm{g}||$ show the $l^2$ norm of the gradient at $\bm{\basalfriction}_k$. Columns titled \#Stokes show the total number of Stokes PDE solves performed in each Newton iteration. This consists of Stokes solves for $u$ performed during the linesearch going from $\bm{\basalfriction}_{k-1}$ to $\bm{\basalfriction}_k$, plus one adjoint Stokes solve to compute the gradient at $\bm{\basalfriction}_k$, plus one incremental forward and one incremental adjoint Stokes solve per PCG iteration for solving the Newton system. In the PSF (5) portion of row 3, we write $6+22$ to indicate that 6 Stokes solves were used during the standard course of the iteration, and 22 Stokes solves were used to build the PSF (5) preconditioner. }
	\label{tab:newton_convergence_table}
\end{table}

@@

\begin{figure}
	\begin{subfigure}{0.59\textwidth}
		\begin{center}
			\begin{tikzpicture}[baseline, trim axis right]
			\begin{axis}[width=0.99\linewidth, height=0.8\linewidth,
			ymode=log,
			xmin=-50,
			xmax=1450,
			ymax=1.e4,
			xlabel={$k$, generalized eigenvalue \#},
			ylabel={$\lambda_{k}$},
			xtick={0, 300, 600, 900, 1200},
			%legend pos=north east,
			legend style={font=\small},
			ylabel near ticks, ylabel shift={-6pt},
			axis x line*=bottom,
			axis y line*=left,
			legend style={at={(0.95,1)},anchor=north east},
			title={Generalized eigenvalues, $\bm{H} \bm{u}_{k}=\lambda_{k} \bm{\preconditioner} \bm{u}_{k}$}
			]
			\addlegendentry{REG}
			\addplot[dotted, line width=0.5mm] table[x=k, y=reg] {generalizedEigenvalues.dat};
			\addlegendentry{PSF (1)}
			\addplot[blue, line width=0.5mm] table[x=k, y=psf1] {generalizedEigenvalues.dat};
			\addlegendentry{PSF (5)}
			\addplot[red, line width=0.5mm] table[x=k, y=psf5] {generalizedEigenvalues.dat};
			\addlegendentry{PSF (25)}
			\addplot[purple, line width=0.5mm] table[x=k, y=psf25] {generalizedEigenvalues.dat};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{}
		\label{fig:generalized_eigenvalues}
	\end{subfigure}
	\begin{subfigure}{0.39\textwidth}
		\begin{center}
			\begingroup
			\setlength{\tabcolsep}{4pt}
			\renewcommand{\arraystretch}{1.25}
			\begin{tabular}{c| c c}
				& $\kappa(\bm{\preconditioner}^{-1} \bm{H})$ &	$\kappa(\bm{\preconditioner}^{-1} \bm{H}_{\text{gn}})$ \\
				\hline 
				REG        & 7.02e+3 & 7.02e+3 \\
				NONE       & 8.71e+2 & 9.22e+2 \\
				PSF $(1)$  & 1.75e+1 & 1.76e+1 \\
				PSF $(5)$  & 1.39e+1 & 1.39e+1 \\
				PSF $(25)$ & 1.03e+1 & 1.06e+1
			\end{tabular}
			\endgroup
		\end{center}
		\caption{}
		\label{fig:condition_number}
	\end{subfigure}
	\caption{Left (\ref{fig:generalized_eigenvalues}): eigenvalues for generalized eigenvalue problem $\bm{H} \bm{u}_k = \lambda_k \bm{\preconditioner} \bm{u}_k$. Here $\bm{H}$ is the Hessian and $\bm{\preconditioner}$ is the PSF approximation constructed using the Gauss-Newton Hessian, $\bm{H}_\text{gn}$, for 1, 5, and 25 batches (PSF (1), PSF (5), and PSF (25), respectively), or the regularization Hessian (REG). Right (\ref{fig:condition_number}): condition number, $\kappa$, for $\bm{\preconditioner}^{-1} \bm{H}$ and $\bm{\preconditioner}^{-1} \bm{H}_\text{gn}$ for these different preconditioners, and no preconditioner (NONE). All operators are evaluated at the $\bm{\basalfriction}$ that solves the inverse problem for $\gamma=7.3 \times 10^3$, as determined by the Morozov discrepancy principle.}
\end{figure}
